Experience Report: Haskell in Computational Biology

### 2012-03-08 mrfy_experience_report.tex
a few more edits. still working through NR's comments.

Proteins are cellular machinery. They interact with one another and with other 
molecules to carry out the functions of living cells: metabolism, regulation, 
signalling, and more.
A~protein's function is determined by its structure, 
and its structure is determined by the sequence of amino acids that
form the protein.
\remark{Lost: there are only 20-odd amino acids}
The amino-acid sequence is ultimately determined by a sequence of
nucleic acids in DNA, which we call a gene.
Given a genetic sequence, biologists wish to know the cellular
function of the protein that gene codes for.
The best known method of discovering such function is
to find other proteins of 
similar structure, which likely share similar function.
This problem is called 
\textit{homology detection}.

Computational biologists detect homologies by building 
algorithms which, given an amino-acid sequence,\remark{This sequence
  (the sequence ``under test'') needs a name.  (Urk! I see below it is
  called the ``query sequence.''  This name needs to come in earlier.) ---NR}
find known proteins of
similar structure.
When the similar proteins are formed from amino-acid sequences that
are nearly identical to the sequence under test, the homologies can be
detected by  
simple string-matching algorithms.\remark{Can we just drop the
  sentence about string matching?}
When the similar proteins are formed from amino-acid sequences that
are not too far in edit distance from the sequence under test, the homologies can be
detected by
a family of algorithms called 
\textit{hidden Markov models}~\cite{hmmer}.\remark{citation needed?}
But in real biological systems,
proteins with similar structure and function may be formed from significantly 
different amino-acid sequences, which are not close in edit distance.
Our~research contribution---the MRFy algorithm---can detect homologies
in amino-acid sequences that are only distantly related.
%
%We will explain an 
%algorithm for detecting reasonably similar sequences, and then move on to 
%explain MRFy, our novel approach for detecting more distantly related sequences 
%for proteins that share similar structure and function.

@@

Homology-detection software most often used in one of two ways:
to test a hypothesis about 
the function of a single, newly discovered protein, or 
to compare every protein in a genome against a library of known protein 
structures.
\remark{Drop this sentence? This latter case demands particularly high-performance software.}
In~both cases, the software is ``trained'' \remark{do we want to talk ``Training''?}
on a group of proteins that share function and structure.
These proteins are identified by a biologist, who puts
their amino-acid sequences into a correspondence, also called an
\emph{alignment}.
\remark{What does the alignment signify?  Biological significance?
  Computational significance?}
An~alignment may be represented as a matrix
(Figure~\ref{alignment}) 
in which each row corresponds to the amino-acid sequence of a protein,
and each column groups amino acids that play similar roles in
different proteins.
A~column
may contain gaps, which indicate
indicating certain protein have no amino acid participating in that
position.
\remark{The word ``position'' hasn't been defined, and can we get away
  without ``residue''?  (If~we need ``residue,'' let's give it a
  proper introduction.}
If~a column contains just one or two gaps, the most likely explanation
is that in one or two proteins, mutations in the underlying DNA
deleted an amino acid from that position (\figref{alignment}).
If~a column contains \emph{mostly} gaps, 
the most likely explanation
is that in one or two proteins, mutations in the underlying DNA
\emph{inserted} an amino acid in that position (\figref{alignment}).

The protein alignment is used to train a \emph{hidden-Markov model}.
\remark{I've never seen this hypenated, in ANY literature.}
A~hidden-Markov model is a probabilistic finite-state machine which 
can assign a probability to any amino-acid sequence under test.
\remark{What is the biological or computational meaning of the
  probability?}
Proteins whose sequences have higher probabilities are more likely to
be structurally related to the proteins in the alignment.
\remark{Should we say ``homologous'' rather than  ``structurally
  related''?
  If~so, where should this jargon be introduced?}
The states of the hidden-Markov model are determined by the alignment:
\remark{I've tried for lots of parallel structure.
Somebody let me know if any of this is right.  ---NR}
\begin{itemize}
\item
For a given column~$C$ of the alignment, the hidden-Markov model has a
\emph{match state}.
The match state determines the probability that the sequence under
test has a particular amino acid in column~$C$.
The match state assigns high probability to the amino acids that
appear in column~$C$, and a low but nonzero probability to other amino
acids.
\item
For every column of the alignment, the hidden-Markov model has a
\emph{deletion state}.
The deletion state determines the probability that the sequence under
test has \emph{no} amino acid  in column~$C$, i.e., that it has a gap.
The deletion state has high probability if there are lots of gaps in
column~$C$,
and low probability if there are few gaps in column~$C$.
\item
Between every adjacent pair of columns in the alignment,\remark{This isn't right}
the hidden-Markov model has an
\emph{insertion state}.\remark{concept of consensus columns is what's missing}
The insertion state determines the probability that the sequence under
test has an ``extra'' amino acid that does not correspond to any
column in the alignment.
The insertion state assigns high probability to \emph{[some amino
    acids]}, and low probability to \emph{[other amino acids]}.
\end{itemize}
%%  A position that has residues in a majority of the sequences becomes 
%%  a \textit{match} state in the hidden Markov model; this position may see a 
%%  variety of amino acids due to evolutionary substitution, but it still 
%%  corresponds to a position that a homologous protein should most likely fill.
The hidden-Markov model also has distinguished \emph{begin} and \emph{end} states.
\figref{plan7} shows a hidden-Markov model derived from an alignment
with just four columns.
\emph{[I'm sure I don't have the proper explanation of the probabilities.]}
\remark{LOST: Each edge in the 
graph has a weight which is the probability of the associated state transition.}
Each column in the original alignment corresponds to a \emph{node},
which is a triplet holding an insertion state, a deletion state, and a
match state.
What looks like an 
intimidating geometry is thus completely explained by the fact that we have 
insertions, deletions, and substitutions due to evolution.
\remark{The evolution seems like a distraction to me here, and we hit
  it above.  Can we simply drop this sentence?}
\emph{[HOW ABOUT SOME HASKELL TYPE DEFINITIONS AT OR NEAR THIS POINT?
    I THINK WE'RE READY.]}

>>>>

Proteins are cellular machinery. They interact with one another and with other 
molecules to carry out the functions of living cells: metabolism, regulation, 
signalling, and more.
A~protein's function is determined by its structure, 
and its structure is determined by the sequence of amino acids that
form the protein.
\remark{Lost: there are only 20-odd amino acids}
The amino-acid sequence is ultimately determined by a sequence of
nucleic acids in DNA, which we call a gene.
Given a genetic sequence (the \textit{query sequence}), biologists wish to know the cellular
function of the protein that gene codes for.
The best known method of discovering such function is
to find other proteins of 
similar structure, which likely share similar function.
Proteins that share structure and function are \textit{homologous}, and thus
the process of identifying similar proteins to a query sequence is called 
\textit{homology detection}.

Computational biologists detect homologies by building 
algorithms which, given an amino-acid query sequence,
find known proteins of similar structure.
When the similar proteins are formed from amino-acid sequences that
are nearly identical to the sequence under test, the homologies can be
detected by  
simple string-matching algorithms.\remark{Can we just drop the
  sentence about string matching?}
When the similar proteins are formed from amino-acid sequences that
are not too far in edit distance from the sequence under test, the homologies can be
detected by
a family of algorithms called 
\textit{hidden Markov models}~\cite{hmmer}.
But in real biological systems,
proteins with similar structure and function may be formed from significantly 
different amino-acid sequences, which are not close in edit distance.
Our~research contribution---the MRFy algorithm---can detect homologies
in amino-acid sequences that are only distantly related.
%
%We will explain an 
%algorithm for detecting reasonably similar sequences, and then move on to 
%explain MRFy, our novel approach for detecting more distantly related sequences 
%for proteins that share similar structure and function.

@@

Homology-detection software most often used in one of two ways:
to test a hypothesis about 
the function of a single, newly discovered protein, or 
to compare every protein in a genome against a library of known protein 
structures.
The software is ``trained'' \remark{do we want to talk ``Training''?}
on a group of proteins that share function and structure.
These proteins are identified by a biologist, who puts
their amino-acid sequences into a correspondence, also called an
\emph{alignment}. This alignment signifies structural and evolutionary
correspondence among individual amino acids in a set of homologous proteins.
An~alignment may be represented as a matrix
(Figure~\ref{alignment}) 
in which each row corresponds to the amino-acid sequence of a protein,
and each column groups amino acids that play similar roles in
different proteins.
A protein may have a gap in a column of the alignment, which indicates
that during the course of evolution, either that protein lost that amino
acid, or other proteins gained an amino acid.
If~a column contains just one or two gaps, the most likely explanation
is that in one or two proteins, mutations in the underlying DNA
deleted an amino acid from that position (\figref{alignment}).
If~a column contains \emph{mostly} gaps, 
the most likely explanation
is that in one or two proteins, mutations in the underlying DNA
\emph{inserted} an amino acid in that position (\figref{alignment}).

The protein alignment is used to train a \emph{hidden-Markov model}.
\remark{I've never seen this hypenated, in ANY literature.}
A~hidden-Markov model is a probabilistic finite-state machine which 
can assign a probability to any amino-acid sequence under test.
\remark{What is the biological or computational meaning of the
  probability?}
Proteins whose sequences have higher probabilities are more likely to
be homologous to the proteins in the alignment.
\remark{Should we say ``homologous'' rather than  ``structurally
  related''?
  If~so, where should this jargon be introduced?}
The states of the hidden-Markov model are determined by the alignment:
\remark{I've tried for lots of parallel structure.
Somebody let me know if any of this is right.  ---NR}
\begin{itemize}
\item
If a given column~$j$ of the alignment contains amino acids in a majority
of the protein sequences, it is considered a \textit{consensus column}.
\item
For a given column~$j$ of the alignment, the hidden-Markov model has a
\emph{match state}.
The match state determines the probability that the sequence under
test has a particular amino acid in column~$j$.
The match state assigns high probability to the amino acids that
appear in column~$j$, and a low but nonzero probability to other amino
acids.
The match state is more probable if it corresponds to a consensus column.
\item
For every column of the alignment, the hidden-Markov model has a
\emph{deletion state}.
The deletion state determines the probability that the sequence under
test has \emph{no} amino acid  in column~$j$, i.e., that it has a gap.
The deletion state has high probability if there are lots of gaps in
column~$j$, and low probability if there are few gaps in column~$j$.
Thus, the deletion state is more likely in non-consensus columns.
\item
For every column of the alignment, the hidden-Markov model has an
\emph{insertion state}.
The insertion state determines the probability that a query sequence
has an ``extra'' amino acid that does not correspond to any
consensus column in the alignment.
Thus, the insertion state is more likely in non-consensus columns.
The insertion state assigns high probability to amino acids that are
frequently observed among all proteins, and low probability to amino acids
that are rarely observed.
\end{itemize}
%%  A position that has residues in a majority of the sequences becomes 
%%  a \textit{match} state in the hidden Markov model; this position may see a 
%%  variety of amino acids due to evolutionary substitution, but it still 
%%  corresponds to a position that a homologous protein should most likely fill.
The hidden-Markov model also has distinguished \emph{begin} and \emph{end} states.
\figref{plan7} shows a hidden-Markov model derived from an alignment
with just four columns.
\emph{[I'm sure I don't have the proper explanation of the probabilities.]}
\remark{LOST: Each edge in the 
graph has a weight which is the probability of the associated state transition.}
Each column in the original alignment corresponds to a \emph{node},
which is a triplet holding an insertion state, a deletion state, and a
match state.
What looks like an 
intimidating geometry is thus completely explained by the fact that we have 
insertions, deletions, and substitutions due to evolution.
\remark{The evolution seems like a distraction to me here, and we hit
  it above.  Can we simply drop this sentence?}
\emph{[HOW ABOUT SOME HASKELL TYPE DEFINITIONS AT OR NEAR THIS POINT?
    I THINK WE'RE READY.]}
